HAMdetector: A Bayesian regression model that integrates information to detect HLA-associated mutations

### 2021-03-25 docs/main.tex
previously uncommitted changes (cannot remember what they were)

\access{Advance Access Publication Date: Day Month Year}
\appnotes{Manuscript Category}
\usepackage{soul}
\usepackage[paper=b4paper, layout=b4paper]{geometry}
\usepackage{float}
% \usepackage[section]{placeins}

@@

\maketitle
\sloppy
\section{Introduction}
\subsection{The HLA system}

@@

\begin{figure}
 \includegraphics[width=1.0\linewidth]{plots/phylogeny_calibration.pdf}  
  \caption{Calibration plot for the HBV core protein.}
  \label{fig:calibration}
\end{figure}

@@

The predictions of all tested models are strongly enriched in epitopes over baselines, irrespective of dataset (Fig.~\ref{fig:comparison} for HBe protein and Supplementary Figures ...-... for other datasets). For the best ranked HAMs, Fisher's exact test performs about as well as the HAMdetector backbone logistic regression model (model~1 in Fig.~\ref{fig:comparison}). Each of the following three model stages of HAMdetector increases HAM enrichment further, independently of the dataset.
The horseshoe prior alone (model~2) is a drastic improvement over model~1, even though it does not include any external information. The logistic regression model with horseshoe prior works roughly as well as the Phylogenetic Dependency Network \cite{Carlson2008}, which includes much more information. Note that the line for the Phylogenetic Dependency Network is shorter because it only outputs associations with a q-value lower than 0.2 by default. Model~3 with its additional inclusion of phylogeny has higher enrichment than model~2, and finally, the full model~4 with the inclusion of epitope prediction leads to a further improvement. Note that model~4 only uses epitope \textit{prediction} software and does not use any information of experimentally confirmed epitopes. These are here only used for model evaluation.

\begin{figure}
  \centering
  \includegraphics[width=0.8\linewidth]{plots/HBe.pdf}
  \caption{HAM enrichment plot for HBe protein: number $N_e$ of associations inside the boundary of known epitopes vs.\ rank $r$. D: Phylogenetic Dependency Network; F: Fisher's exact test; 1: simple logistic regression model with broad Student-t priors; 2: logistic regression model with horseshoe prior; 3: logistic regression model with horseshoe prior and phylogeny; 4: full model with epitope prediction. Unannotated gray lines at the bottom of the graph are HAM enrichment curves for random permutations of the list of HLA allele - replacement pairs and act as baselines.}
  \label{fig:comparison}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=0.8\linewidth]{plots/without_horseshoe.pdf}
  \includegraphics[width=0.8\linewidth]{plots/with_horseshoe.pdf}
  \caption{Marginal posterior distributions of regression coefficients for the association of replacement 11D of the HIV integrase with HLA alleles A*01 and B*44. Top half: inferred with logistic regression model, bottom half: inferred with logistic regression with sparsifying horseshoe prior.}
  \label{fig:horseshoe-comparison}
\end{figure}

@@

\begin{table}
  \renewcommand{\arraystretch}{1.3}
  \centering
  \caption{ELPD changes as HAMdetector components are added. Data computed for HBe protein of the HBV dataset. All differences in ELPD are larger than several times the estimated standard error (column \(\text{se}_\text{diff}\)), indicating that models that include more information have better predictive performance.}
  \vspace{0.5cm}
  \begin{tabular}{l|r|rll}
  \multicolumn{1}{l|}{} & \multicolumn{1}{c|}{\(\text{ELPD}_\text{diff}\)} & \multicolumn{1}{c}{\(\text{se}_\text{diff}\)} &  &  \\ \cline{1-3}
  logistic regression (baseline) & 0.0  & 0.0 &  &  \\
  + horseshoe prior     & 949.8  & 65.2 &  &  \\
  + phylogeny & 4440.9 & 94.4 &  &  \\
  + epitope prediction & 63.1   & 18.9 &  & 
  \end{tabular}
  \label{loo}
\end{table}

@@

\subsection{HAMs in HDV as test case}
The Hepatitis D Virus (HDV) dataset \citep{Karimzadeh2019} is an excellent test case: we have (1) a set of paired HDV sequences and patient HLA alleles, (2) HAM predictions by Fisher's exact test as implemented in SeqFeatR \citep{Budeus2016}, and (3) experimental tests of the predicted HAMs for immune escape activity (IFN-\(\gamma\) production assays, \citet{Karimzadeh2019}). This allows us to see whether HAMdetector decreases the false positive rate in comparison to the simpler Fisher's exact test, and we can make \emph{bona fide} predictions on previously undetected HAMs.
We have 15 HAMs predicted in HDV by Fisher's exact test at significance level $5\times 10^{-3}$ (Table~\ref{tab:false-positives}) as published \citep{Karimzadeh2019}. The corresponding p-values  have no clear relation to experimental confirmation, i.e.\ p-values for confirmed HAMs are not generally lower than those of non-confirmed ones.

For HAMdetector, we use in Table~\ref{tab:false-positives} the posterior probability of a positive regression coefficient ($P(\beta_{jk}>0)$ as measure for the confidence in having detected a HAM. HAMs with strong support have a posterior probability close to 1, associations with no support a probability close to 0.5 (corresponding to a regression coefficient centered around 0). The five predicted HAMs with top posterior probabilities (all $\ge 0.90$) have all been experimentally confirmed. There is only one outlier with posterior probability 0.75 (P89T and B*37).

HAMdetector shows strong support for 15 replacement - allele pairs that have previously not been identified. All of them exceed the significance level of 
$5\times 10^{-3}$ used in~\citet{Karimzadeh2019}. This shows that HAMdetector is able to identify possible HAMs that would not be detected otherwise. The most striking example for this is K43R - A*02, which using Fisher's exact test has a p-value of 0.22. While this replacement has not been experimentally verified in a IFN-\(\gamma\) production assay yet, it lies within a known A*02 epitope and thus it is highly likely that K43R conveys some form of immune escape.

\begin{table}
  \caption{List of HAMs predicted Fisher's exact test (FET). In the last column a ``+'' marks experimental confirmation and a ``-'' a failed confirmation. Column ``post.prob.'' shows the corresponding posterior probabilities for positive associations computed with HAMdetector. \emph{Bona fide} predictions are separated by a horizontal line. These are replacement - allele pairs we expect to to be verified in future experimental tests. The replacement annotated with a (+) lies within a known A*02 epitope.}
  % \vspace{0.5cm}
  \begin{tabular}{c|c|c|c|c}
  replacement & allele  & p-value (FET) & post.~prob. & confirmed \\
  \hline
  S170N       & B*15 & $3\cdot10^{-8}$      & 0.99                                & +                        \\
  D101E       & B*37 & 0.0002                        & 0.96                                & +                        \\
  R105K       & B*27 & 0.0011                        & 0.93                                & +                        \\
  R139K       & B*41 & 0.0034                        & 0.92                                & +                        \\
  E47D        & B*18 & 0.0027                        & 0.90                                & +                        \\
  D33E        & B*13 & 0.0001                        & 0.86                                & -                        \\
  T134A       & A*68 & 0.0045                        & 0.82                                & -                        \\
  K43R        & B*13 & 0.0021                        & 0.77                                & -                        \\
  P89T        & B*37 & 0.0011                        & 0.75                                & +                        \\
  D47E        & A*30 & 0.0010                        & 0.76                                & -                        \\
  K113R       & B*13 & 0.0043                        & 0.76                                & -                        \\
  A107T       & B*14 & 0.0028                        & 0.70                                & -                        \\
  P49L        & A*30 & 0.0031                        & 0.63                                & -                        \\
  Q100L       & B*13 & 0.0018                        & 0.60                                & -                        \\
  D96E        & B*13 & 0.0035                        & 0.51                                & -  \\\hline
  E46D & A*02 & 0.0054 & 0.97 & ? \\
  V81I & A*68 & 0.0073 & 0.97 & ? \\
  K113N & B*08 & 0.0063 & 0.96 & ? \\
  A71T & B*41 & 0.0065 & 0.96 & ? \\
  L188I & A*68 & 0.0632 & 0.94 & ? \\
  T95S & A*01 & 0.0285 & 0.93 & ? \\
  D33E & A*03 & 0.0226 & 0.93 & ? \\
  P49L & B*13 & 0.0035 & 0.92 & ? \\
  A74S & A*68 & 0.0123 & 0.91 & ? \\
  E29D & B*44 & 0.0559 & 0.91 & ? \\
  D46E & B*57 & 0.0190 & 0.91 & ? \\
  R88K & A*68 & 0.0123 & 0.91 & ? \\
  T149P & B*52 & 0.0281 & 0.91 & ? \\
  K43R & A*02 & 0.2158 & 0.90 & (+) \\
  N22S & B*08 & 0.0405 & 0.90 & ? \\
  \end{tabular}
  \label{tab:false-positives}
 \end{table}

@@

 Table~\ref{tab:false-positives} shows that for 3 of the false positives identified with Fisher's exact test, HAMdetector identifies associations with the same replacement but a different allele (P49L - B*13 instead of P49L - A*30; K43R - A*02 instead of K43R - B*13; and D33E - B*13 instead of D33E - A*03).
 One possible explanation for this observation is HLA-Linkage disequilibrium and accordingly correlation between HLA alleles in the dataset due to random sampling variation.
Indeed, out of the 12 times P49L is observed in sequences annotated with A*30, B*13 is also present in 5 of those cases (Spearman's rank correlation coefficient \(\rho = 0.5\)). A similar observation can be done for K43R and D33E, albeit the correlation between the respective alleles is much weaker.

\begin{figure}
  \includegraphics[width=1.1\linewidth]{plots/P49L_combined.pdf}
  \caption{Marginal posteriors for the regression coefficients of A*30 and B*13 for replacement P49L. The association between P49L and A*30 identified by Fisher's exact test turned out to be a false positive.}
  \label{fig:linkage}
\end{figure}

Figure~\ref{fig:linkage} shows regression coefficients of the HLA alleles A*30 and B*13 for replacement P49L. When using a simple logistic regression model without any additions (model~1), both A*30 and B*13 show medium evidence of being associated with replacement P49L. However, when also taking phylogeny into account and using a sparsity-promoting prior (model~3), both regression coefficients get shrunk to 0 because the association is not strongly enough identified by the data alone.
Using epitope prediction as an additional source of information (model~4) allows to disentangle the association of the correlated alleles with P49L and identify B*13 as associated with P49L. The false-positive association between P49L and A*30 remains being shrunk towards 0.  
\section{Discussion}

@@

%\section*{Acknowledgements}
% \FloatBarrier
\section*{Funding}

>>>>

\access{Advance Access Publication Date: Day Month Year}
\appnotes{Manuscript Category}
\usepackage{soul}
\usepackage[paper=b4paper, layout=b4paper]{geometry}

@@

\maketitle

\section{Introduction}
\subsection{The HLA system}

@@

\begin{figure}[!ht]
 \includegraphics[width=1.0\linewidth]{plots/phylogeny_calibration.pdf}  
  \caption{Calibration plot for the HBV core protein.}
  \label{fig:calibration}
\end{figure}

@@

The predictions of all tested models are strongly enriched in epitopes over baselines, irrespective of dataset (Fig.~\ref{fig:comparison} for HBe protein and Supplementary Figures \hl{XXX-YYY} for other datasets). For the best ranked HAMs, Fisher's exact test performs about as well as the HAMdetector backbone logistic regression model (model 1 in Fig.~\ref{fig:comparison}). Each of the following three model stages of HAMdetector increases HAM enrichment further, independently of the dataset.
The horseshoe prior alone (model 2) is a drastic improvement over model 1, even though it does not include any external information. The logistic regression model with horseshoe prior works roughly as well as the Phylogenetic Dependency Network \cite{Carlson2008}, which includes much more information. Note that the line for the Phylogenetic Dependency Network is shorter because it only outputs associations with a q-value lower than 0.2 by default. Model 3 with its additional inclusion of phylogeny has higher enrichment than model 2, and finally, the full model 4 with the inclusion of epitope prediction leads to a further improvement. Note that model 4 only uses epitope \textit{prediction} software and does not use any information of experimentally confirmed epitopes. These are here only used for model evaluation.

\begin{figure}[ht!]
  \includegraphics[width=1\linewidth]{plots/comparison.pdf}
  \caption{HAM enrichment plot for HBe protein: number $N_e$ of associations inside the boundary of known epitopes vs.\ rank $r$. PhyloD: Phylogenetic Dependency Network; model 1: simple logistic regression model with broad Student-t priors; model 2: logistic regression model with horseshoe prior; model 3: logistic regression model with horseshoe prior and phylogeny; model 4: full model with epitope prediction. Gray lines are HAM enrichment curves for random permutations of the list of HLA allele - replacement pairs and act as baselines.}
  \label{fig:comparison}
\end{figure}

@@

\begin{figure}
  \includegraphics[width=1\linewidth]{plots/without_horseshoe.pdf}
  \includegraphics[width=1\linewidth]{plots/with_horseshoe.pdf}
  \caption{Marginal posterior distributions of regression coefficients for the association of replacement 11D of the HIV integrase with HLA alleles A*01 and B*44. Top half: inferred with logistic regression model, bottom half: inferred with logistic regression with sparsifying horseshoe prior.}
  \label{fig:horseshoe-comparison}
\end{figure}

@@

\begin{table}[h!]
  \renewcommand{\arraystretch}{1.3}
  \centering
  \caption{ELPD changes as HAMdetector components are added. Data computed for HBe protein of the HBV dataset. All differences in ELPD are larger than several times the estimated standard error (column \(\text{se}_\text{diff}\)), indicating that models that include more information have better predictive performance.}
  \vspace{0.5cm}
  \begin{tabular}{l|r|rll}
  \multicolumn{1}{l|}{} & \multicolumn{1}{c|}{\(\text{ELPD}_\text{diff}\)} & \multicolumn{1}{c}{\(\text{se}_\text{diff}\)} &  &  \\ \cline{1-3}
  logistic regression (baseline) & 0.0  & 0.0 &  &  \\
  + horseshoe prior     & 949.8  & 65.2 &  &  \\
  + phylogeny & 4440.9 & 94.4 &  &  \\
  + epitope prediction & 63.1   & 18.9 &  & 
  \end{tabular}
  \label{loo}
\end{table}

@@

\subsection{HAMs in HDV as test case}
The Hepatitis D Virus (HDV) dataset \citep{Karimzadeh2019} is an excellent test case: we have (1) a set of paired HDV sequences and patient HLA alleles, (2) HAM predictions by Fisher's exact test as implemented in SeqFeatR \citep{Budeus2016}, and (3) experimental tests of the predicted HAMs for immune escape activity (IFN-\(\gamma\) production assays, \citet{Karimzadeh2019}). This allows us to see whether HAMdetector decreases the false positive rate in comparison to the simpler Fisher's exact test, and we can make \emph{bona fide} predictions on previously undetected HAMs. \hl{DaHo: zu \emph{bona fide}: können wir dazu etwas sagen? Ich erinnere mich, dass Du noch eine Reihe guter HAMs vorhergesagt hattest.}

We have 15 HAMs predicted in HDV by Fisher's exact test at significance level $5\times 10^{-3}$ (Table~\ref{tab:false-positives}) as published \citep{Karimzadeh2019}. The corresponding p-values  have no clear relation to experimental confirmation, i.e.\ p-values for confirmed HAMs are not generally lower than those of non-confirmed ones.

For HAMdetector, we use in Table~\ref{tab:false-positives} the posterior probability of a positive regression coefficient ($P(\beta_{jk}>0)$ as measure for the confidence in having detected a HAM. HAMs with strong support have a posterior probability close to 1, associations with no support a probability close to 0.5 (corresponding to a regression coefficient centered around 0). The five predicted HAMs with top posterior probabilities (all $\ge 0.89$) have all been experimentally confirmed. There is only one outlier with posterior probability 0.75 (P89T and B*37).

\begin{table}[h!]
  \caption{List of HAMs predicted Fisher's exact test (FET). In the last column a ``+'' marks experimental confirmation and a ``-'' a failed confirmation. Column ``post.prob.'' shows the corresponding posterior probabilities for positive associations computed with HAMdetector.}
  \vspace{0.5cm}
  \begin{tabular}{c|c|c|c|c}
  replacement & allele  & p-value (FET) & post.~prob. & confirmed \\
  \hline
  S170N       & B*15 & $3\cdot10^{-8}$      & 0.99                                & +                        \\
  D101E       & B*37 & 0.0002                        & 0.96                                & +                        \\
  R105K       & B*27 & 0.0011                        & 0.93                                & +                        \\
  R139K       & B*41 & 0.0034                        & 0.92                                & +                        \\
  E47D        & B*18 & 0.0027                        & 0.90                                & +                        \\
  D33E        & B*13 & 0.0001                        & 0.86                                & -                        \\
  T134A       & A*68 & 0.0045                        & 0.82                                & -                        \\
  K43R        & B*13 & 0.0021                        & 0.77                                & -                        \\
  P89T        & B*37 & 0.0011                        & 0.75                                & +                        \\
  D47E        & A*30 & 0.0010                        & 0.76                                & -                        \\
  K113R       & B*13 & 0.0043                        & 0.76                                & -                        \\
  A107T       & B*14 & 0.0028                        & 0.70                                & -                        \\
  P49L        & A*30 & 0.0031                        & 0.63                                & -                        \\
  Q100L       & B*13 & 0.0018                        & 0.60                                & -                        \\
  D96E        & B*13 & 0.0035                        & 0.51                                & -  \\\hline
  E46D & A*02 & 0.0054 & 0.97 & ? \\
  V81I & A*68 & 0.0073 & 0.97 & ? \\
  K113N & B*08 & 0.0063 & 0.96 & ? \\
  A71T & B*41 & 0.0065 & 0.96 & ? \\
  L188I & A*68 & 0.0632 & 0.94 & ? \\
  T95S & A*01 & 0.0285 & 0.93 & ? \\
  D33E & A*03 & 0.0226 & 0.93 & ? \\
  P49L & B*13 & 0.0035 & 0.92 & ? \\
  A74S & A*68 & 0.0123 & 0.91 & ? \\
  E29D & B*44 & 0.0559 & 0.91 & ? \\
  D46E & B*57 & 0.0190 & 0.91 & ? \\
  R88K & A*68 & 0.0123 & 0.91 & ? \\
  T149P & B*52 & 0.0281 & 0.91 & ? \\
  K43R & A*02 & 0.2158 & 0.90 & ? \\
  N22S & B*08 & 0.0405 & 0.90 & ? \\
  \end{tabular}
  \label{tab:false-positives}
 \end{table}

@@

 \hl{DaHo: bitte Beispiel mit Linkage Disequilibrium zeigen}

\section{Discussion}

@@

%\section*{Acknowledgements}

\section*{Funding}
